The contact value approximation for the pair distribution function of the inhomogeneous hard-sphere fluid

### 2014-05-16 papers/pair-correlation/paper.tex
added equation equation for previously used pair correlation function Tested on crystal. Took 0:38.82.

\section{Previous approximations to the pair distribution function}
There are several analytic approximations for the inhomogeneous pair
distribution function, which extend the radial distribution function
to inhomogeneous scenarios.  Work on these approximations date back to
Toxvaerd \cite{toxvaerd1973statistical} and have been tested
rigorously for both functional and non-functional based theories
\cite{kalos1977structure,osborn1980monotonic,mccoy1981comparison}.
The approximations differ in what density is used when evaluating the
radial distribution function $g(r,\eta)$ or in how radial
distributions evaluated at different densities are combined to yield
an effective function \cite{toxvaerd1973statistical}.

One common approximation \cite{gloor2007prediction, gross2009density}
assumes a mean-density approximation for the pair distribution
function:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \tfrac{\pi\sigma^3}{6}\tfrac12
  (n(\rr_1)+n(\rr_2))\right)
\end{align}
where $g(r,\eta)$ is the radial distribution function as a function of
radius and packing fraction $\eta$, and $\sigma$ is the hard-sphere
diameter.  This approach,
however, cannot be applied to highly inhomogeneous systems such as a
dense fluid at a solid surface.  These systems exhibit large density
oscillations that lead to mean packing fractions greater than unity,
which cannot occur in the bulk reference system that defines $g(r,
\eta)$.  The above papers restricted themselves to the liquid-vapor
interface, which does not exhibit this pathology.

Non-pathological approaches use an average of the density over some
volume. Fischer and
Methfessel~\cite{fischer1980born,harris1987comment} introduce the
approximation:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is an integral of the density over a spherical volume that
is now used as one of the fundamental measures in Fundamental
Measure Theory~\cite{rosenfeld1989free}:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
Equation~\ref{eq:fischer} is computationally awkward, because it
treats as special the midpoint $\tfrac12(\rr_1+\rr_2)$.  Moreover, the
approach of Fischer and Methfessel is intended to approximate the pair
distribution function only at contact, when the distance between
$\rr_1$ and $\rr_2$ is the hard-sphere diameter.

Sokolowski and Fischer addressed the shortcomings of the theory of
Fischer and Methfessel by modifying this approach to use
averages, centered on the two points $\rr_1$ and $\rr_2$:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12},
  \tfrac12(\bar{n}(\rr_1)+\bar{n}(\rr_2))\right)
  \label{eq:sokolowski}
\end{align}
where their averaged density $\bar{n}(\rr)$ is given by:
\begin{align}
  \bar{n}(\rr) \equiv \frac{3}{4\pi (0.8\sigma)^3}\int n(\rr')\Theta(0.8\sigma - |\rr-\rr'|) d\rr'
\end{align}
is the density averaged over a sphere with diameter
$0.8\sigma$~\cite{sokolowski1992role}.  The value 0.8 in this formula
was arrived at by fitting to Monte Carlo simulation.  Although
Eq.~\ref{eq:sokolowski} has the advantages of only involving density
averages at the points at which the pair distribution function is
desired, it cannot be written as a single-site convolution, since the
convolution kernel depends on both points.

>>>>

\section{Previous approximations to the pair distribution function}
There are several analytic approximations for the inhomogeneous pair
distribution function, which extend the radial distribution function
to inhomogeneous scenarios.  Work on these approximations date back to
Toxvaerd \cite{toxvaerd1973statistical} and have been tested
rigorously for both functional and non-functional based theories
\cite{kalos1977structure,osborn1980monotonic, mccoy1981comparison}.
The approximations differ in what density is used when evaluating the
radial distribution function $g(r,\eta)$ or in how radial
distributions evaluated at different densities are combined to yield
an effective function \cite{toxvaerd1973statistical}.

One common approximation \cite{gloor2007prediction, gross2009density,
  osborn1980monotonic, mccoy1981comparison,barrett2006some} assumes a
mean-density approximation for the pair distribution function:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \tfrac{\pi\sigma^3}{6}\tfrac12
  (n(\rr_1)+n(\rr_2))\right)
\end{align}
where $g(r,\eta)$ is the radial distribution function as a function of
radius and packing fraction $\eta$, and $\sigma$ is the hard-sphere
diameter.

Another approximation uses $g(r,\eta)$ and assumes the density
evaulated at the position between the two points:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \tfrac{\pi\sigma^3}{6}
  n\left(\tfrac12(\rr_1+\rr_2)\right)\right)
\end{align}

The above two approaches cannot be applied to highly inhomogeneous
systems such as a dense fluid at a solid surface.  These systems
exhibit large density oscillations that lead to mean packing fractions
greater than unity, which cannot occur in the bulk reference system
that defines $g(r, \eta)$.  The above papers restricted themselves to
the liquid-vapor interface, which does not exhibit this pathology.

Non-pathological approaches use an average of the density over some
volume. Fischer and Methfessel introduce the
approximation~\cite{fischer1980born,harris1987comment}:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is an integral of the density over a spherical volume that
is now used as one of the fundamental measures in Fundamental
Measure Theory~\cite{rosenfeld1989free}:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
Equation~\ref{eq:fischer} is computationally awkward, because it
treats as special the midpoint $\tfrac12(\rr_1+\rr_2)$.  Moreover, the
approach of Fischer and Methfessel is intended to approximate the pair
distribution function only at contact, when the distance between
$\rr_1$ and $\rr_2$ is the hard-sphere diameter.  \davidsays{Tang does something similar, but maybe different.\cite{tang1991density}}

Sokolowski and Fischer addressed the shortcomings of the theory of
Fischer and Methfessel by modifying this approach to use
averages, centered on the two points $\rr_1$ and $\rr_2$:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12},
  \tfrac12(\bar{n}(\rr_1)+\bar{n}(\rr_2))\right)
  \label{eq:sokolowski}
\end{align}
where their averaged density $\bar{n}(\rr)$ is given by:
\begin{align}
  \bar{n}(\rr) \equiv \frac{3}{4\pi (0.8\sigma)^3}\int n(\rr')\Theta(0.8\sigma - |\rr-\rr'|) d\rr'
\end{align}
is the density averaged over a sphere with diameter
$0.8\sigma$~\cite{sokolowski1992role, wadewitz2000application, winkelmann2001liquid}.  The
value 0.8 in this formula was arrived at by fitting to Monte Carlo
simulation.  Although Eq.~\ref{eq:sokolowski} has the advantages of
only involving density averages at the points at which the pair
distribution function is desired, it cannot be written as a
single-site convolution, since the convolution kernel depends on both
points.
